# Data Manipulation in R

<hr>

### Daniel J. Hocking
### 22 November 2013
### R Working Group

<hr>

R is a great resource and has become the lingua franca for statistics in ecology. R is not the best of languages but it has to big advantages: a large ecology user base and a large, centralized repository of contributed packages (CRAN). R is an incredibly flexible language, possibly to a fault. For example to extract a column from a data frame (we'll call toy) you can do any of the following:

```{r}
a <- c('one', 'two', 'three', 'four')
b <- c(1, 2, 3, 4)
c <- c(1.1, 2.2, 3.3, 4.4)
toy <- data.frame(a, b, c)

str(toy)
summary(toy)
```

```{r}
toy[ , 1]
toy$a
toy[ ,c('a')]
toy[['a']]
```

As you can see, this all produces the same results. In other languages, you can usually only access a portion of the toy in 1 way. While the flexiblity of R can be useful at times, it can cause confusion and creates an extremely steep learning curve. It's difficult to read the code from other people without a much larger vocabularly.

Add a column to a toyframe
```{r}
d <- seq(from = 100, to = 200, length.out = 4)
d
(e <- seq(100, 200, length.out = 4))
e[3]
(e3 <- e[3] + 0.000002)
print(e[3], dig = 10)
print(e3, dig = 10)

toy$d <- d
toy$new <- e # name in the toyframe independent of original object name
toy$f <- rep(NA, times = 4)

toy$d
toy$d <- c('o', 'v', 'e', 'r') # overwrites existing column with the same name
toy$d
```

Now let's work with some real data. It will be bigger and messier than our toy dataset:
```{r}
setwd('/Users/Dan/Documents/Teaching/R_intro/02_Data_Manipulation/')
library(RCurl)
foo <- getURL("https://raw.github.com/djhocking/R_Intro/master/02_Data_Manipulation/Salamander_Demographics.csv", ssl.verifypeer = TRUE)
demo <- read.table(textConnection(foo), header=TRUE, sep=",", na.strings=NA)
# demo <- read.table('Salamander_Demographics.csv', header = TRUE, sep = ",") # alternatively you can download data from GitHub into your working directory and load from the local file

str(demo)
head(demo)
tail(demo)
summary(demo)
```

Let's create a dataframe with just the size data
```{r}
size.vars <- demo[c("svl", "tl", "mass")]
head(size.vars)
```

or maybe we just want the first 5 rows
```{r}
demo5 <- demo[, c(1:5)]
demo5b <- demo[c(1:5)]

head(demo5)
head(demo5b)
```

Delete some variables
```{r}
rm.vars <- names(demo) %in% c("id", "mark", "recap")
newdemo <- demo[!rm.vars]
head(newdemo)

newdemo2 <- demo[c(-1, -3)]
head(newdemo2)

newdemo2$id <- newdemo2$mark <- NULL
head(newdemo2)
```

Select Observations
```{r}
# based on variable values
newdemo <- demo[ which(demo$sex=='Y' & demo$mass > 1.0), ]
str(demo)
str(newdemo)
```

Subset Function
```{r}
newdemo <- subset(demo, sex=="X" & gravid == "Y", select = mass:tl) 
str(newdemo)
head(newdemo)
```

Make table and export as csv file
```{r}
write.table(x = newdemo, file = "Gravid_Female_Demographics.csv", sep = ",")
```

Sorting Data
There is no undo key in R. If you write over or delete an object or column, it's gone. Similarly, if you sort you can't unsort. I like to have a primary key (line number) so can always return to orginal order.
```{r}
demo$key <- seq(1, length(demo$svl))
head(demo, n = 10)
demo <- demo[order(demo$mass), ] # if don't want to create extra dataframes
demo.sort <- demo[order(demo$mass), ] # alt create new sorted dataframe
head(demo, n = 10)

demo.sort <- demo[order(demo$sex, demo$svl), ]
head(demo.sort, n = 10)
```

### On your own:
Add column of random numbers from your favorite distribution (hint ?rnorm) and sort by tl and then that column from largest to smallest.


### Further exploration:
built-in (base) functions:
by
aggregate

Packages to try:
plyr
reshape2



