# R Plotting Basics

### Daniel J. Hocking
### 09 December 2013
### UNH R Working Group

Set working directory and load data
```{r}
setwd('/Users/Dan/Documents/Teaching/R_intro/03_Plotting_Basics/')

library(RCurl)
foo <- getURL("https://raw.github.com/djhocking/R_Intro/master/Data/Salamander_Demographics.csv", ssl.verifypeer = TRUE)
demo <- read.table(textConnection(foo), header = TRUE, sep = ",", na.strings = NA)
# demo <- read.table('Salamander_Demographics.csv', header = TRUE, sep = ",") # alternatively you can download data from GitHub into your working directory and load from the local file

str(demo)
head(demo)
tail(demo)
summary(demo)
```

Scatterplot
```{r}
plot(demo$svl, demo$mass, xlab = "Snout-vent length (mm)", ylab = "Mass (g)")
```

Add different symbols for males, females, juveniles (UI)
```{r}
plot(demo$svl, demo$mass, xlab = "Snout-vent length (mm)", ylab = "Mass (g)", pch=as.integer(demo$sex))
```

Add colors for males, females, juveniles (UI)
```{r}
plot(demo$svl, demo$mass, xlab = "Snout-vent length (mm)", ylab = "Mass (g)", pch = as.integer(demo$sex), col = as.integer(demo$sex))
```

Add smooth spline (curve through the points)
```{r}
lines(smooth.spline(demo[which(demo$svl != NA & demo$mass != NA)]$svl, demo$mass))

Add legend
```{r}
plot(demo$svl, demo$mass, xlab = "Snout-vent length (mm)", ylab = "Mass (g)", pch = as.integer(demo$sex), col = as.integer(demo$sex))
legend
```







