# R Plotting: ggplot2

### Daniel J. Hocking
### 16 December 2013
### UNH R Working Group

ggplot2 is an R package for graphing that uses a specific grammar to build graphics in a series of layers starting with the raw data. The grammar is based on "Grammar of Graphics" (Wilkinson 2005).   

There is a tremendous amount of information available on ggplot2 online, owing to its popularity. There is also a book describing the package, "ggplot2: Elegant Graphics for Data Analysis" (Wickham 2009). The package creator, Hadley Wickham, provides information on his website: [http://had.co.nz/ggplot2](http://had.co.nz/ggplot2).   


Basic Grammar Components:   
* data: data to visualize
* mapping: aesthetics describing how data are mapped to attributes
* geom: geometric objects printed on plot such as points, lines, polygons 
* stats: optional statistics transformations for binning, counting, and summarizing data
* scale: translates data scales to aesthetic scales like color, size, shape and draws legend and axes
* coord: coordinate system describes how data and mapped to plane of graphic and provides axes and grid lines. Cartesian coordinates, polar coordinates, or map projections
* facet: faceting specifies how to break data into subsets (conditioning, latticing, or trellising)
* fonts, colors, etc. are specied via that theme and not as graphical grammar components   

First, import the salamander demographics data

```{r}
setwd('/Users/Dan/Documents/Teaching/R_intro/04_Plotting_Packages/')

library(RCurl)
foo <- getURL("https://raw.github.com/djhocking/R_Intro/master/Data/Salamander_Demographics.csv", followlocation = TRUE, cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
demo <- read.table(textConnection(foo), header = TRUE, sep = ",", na.strings = NA)
# demo <- read.table('Salamander_Demographics.csv', header = TRUE, sep = ",") # alternatively you can download data from GitHub into your working directory and load from the local file
```

Remember the plot we created previously to visualize mass as a function of snout-vent length with different colors for different sex/age classes (below). We are going to recreate this using ggplot2

```{r}
#par(mar = c(3.5,3,2,1), mgp = c(2,.7,0), tck = -.02)
plot(demo$svl, demo$mass, xlab = "Snout-vent length (mm)", ylab = "Mass (g)", pch = as.integer(demo$sex), col = as.integer(demo$sex))
lines(smooth.spline(demo[which(demo$svl != "NA" & demo$mass != "NA"), ]$svl, demo[which(demo$svl != "NA" & demo$mass != "NA"), ]$mass))
legend(x = 15, y = 1.9, legend = c("Unknown", "Adult", "Juvenile", "Female", "Male"), pch = 1:5, col = 1:5)
```

We will start with a quick plot using the `qplot` function:

```{r}
library(ggplot2)
qplot(svl, mass, data = demo)
```

This looks very similar to our original plot but with a different background and filled points. It also automatically removed missing values and produced a warning to let us know. This is using `geom = "point"` by default to produce a scatterplot because x and y values were supplied. With more ease than with the base `plot` function, we can add different colors and shapes for animals of different sexes and included a legend:

```{r}
qplot(svl, mass, data = demo,   # I only separate rows after the column so you can see what I added
      colour = sex,             # notice the British spelling b/c Hadley is from NZ
      shape = sex)
```

This was much easier than the base plot code! Now let's add the smoothing spline with the geom `smooth`:

```{r}
library(mgcv) # needed for the smooth option with large datasets to do gam
qplot(svl, mass, data = demo[which(demo$sex != "UA"), ],
      colour = sex,
      shape = sex,
      geom = c("point", "smooth"))
```

AFTER the plot is created the non-data part of the graphic can be manipulated by altering the theme. There are built-in themes or you can customize any aspect with your own theme. Themes can also be changed on a global setting with the `theme_set(theme_grey())` funtion or locally with the `qplot(...) + theme_grey()`. 

```{r}
qplot(svl, mass, data = demo[which(demo$sex != "UA"), ]) + geom_smooth(se = FALSE) + theme_bw()
```

Let's make a custom theme

```{r}
theme_eco <- function(base_size = 12, base_family = "Helvetica", legend_position = "right") {
  theme(
    text = element_text(family = base_family, face = "plain", colour = "black", size = base_size),
    panel.background = element_blank(),
    legend.key = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(colour = "black"),
    axis.text.y = element_text(colour = "black"),
    legend.position = legend_position)
  }


p1 <- qplot(svl, mass, data = demo, colour = sex, shape = sex, xlab = "Snout-vent length (mm)", ylab = "Mass (g)")

p1 + theme_eco(legend_position = c(0.1, 0.8))
```

Other quick plots using ggplot include boxplots, histograms, and density plots.

```{r}
library(grid)

plot1 <- qplot(svl, mass, data = demo, geom = "point") + theme_eco()
plot2 <- qplot(svl, data = demo, geom = "histogram") + theme_eco()
plot3 <- qplot(svl, data = demo, geom = "density") + theme_eco()
plot4 <- qplot(sex, svl, data = demo, geom = "boxplot") + theme_eco()

vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))
print(plot1, vp = vplayout(1, 1))
print(plot2, vp = vplayout(1, 2))
print(plot3, vp = vplayout(2, 1))
print(plot4, vp = vplayout(2, 2))
```

Facets are like lattice plots. They easily replicate plots in a grid by a categorical variable.

```{r}

qplot(svl, mass, data = demo) + facet_grid(. ~ sex)

```





